---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---




## Loading and preprocessing the data


Load the necessary libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(lubridate)
library(knitr)
```

Unzip and load the data
```{r message=FALSE, warning=FALSE}
#unzip the data
unzip('activity.zip')

#load the data
data = read_csv('activity.csv') %>%
      clean_names()
```
Check the data

```{r}
str(data)
```


## What is mean total number of steps taken per day?

Calculate the total steps daily
```{r message=FALSE, warning=FALSE}
total_steps_daily = data %>%
      group_by(date) %>%
      summarise(total_steps = sum(steps, na.rm = FALSE))
```
Make a histogram of total number of steps taken each day

```{r warning=FALSE, message=FALSE}
plot_steps = total_steps_daily %>%
      ggplot(aes(total_steps))+
      geom_histogram(fill = 'steelblue')+
      labs(title = 'Histogram of step count',
           y = 'Frequency of count',
           x = '')+
      theme_minimal()

ggsave('plot_steps.png',plot_steps)
plot_steps
```

Calculate and report the median and total number of steps taken per day
```{r}
steps_summary = total_steps_daily%>%
      summarise(mean_steps = mean(total_steps, na.rm = TRUE),
             median_steps = median(total_steps, na.rm = TRUE)) %>%
      as.list() %>%
      unlist

steps_summary
```


## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

Calculate the total steps per interval
```{r message=FALSE, warning=FALSE}
steps_by_interval = data %>%
      group_by(interval) %>%
      summarise(total_steps = sum(steps, na.rm = TRUE))
``` 
Plot the number of steps per interval
```{r}
plot_steps_interval = steps_by_interval %>%
      ggplot(aes(interval, total_steps))+
      geom_line(color = 'steelblue', size = 1.5)+
      labs(title = 'Number of total steps by interval',
           y = 'total steps',
           x = 'interval (mins)')+
      theme_minimal()

ggsave('plot_steps_interval.png',plot_steps_interval)
plot_steps_interval
```
Identify which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
interval_max_count = steps_by_interval %>%
      filter(total_steps == max(total_steps)) %>%
      as.list() %>%
      unlist()

interval_max_count
```

## Imputing missing values

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
map_dbl(data, function(x) sum(is.na(x)))
```

Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

Fill NA values using mean of group
```{r message=FALSE, warning=FALSE}
filled_data = data %>%
      mutate(steps = as.numeric(steps)) %>%
      group_by(date) %>%
      mutate(steps = ifelse(is.na(steps), mean(steps, na.rm = TRUE), steps)) %>%
      mutate(steps = coalesce(steps, 0))
```

Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
filled_data %>%
      write_csv('complete_activity.csv')
```

Compute for the total number of steps per day using the complete data
```{r warning=FALSE, message=FALSE}
total_steps_daily_complete =  filled_data %>%
      group_by(date) %>%
      summarise(total_steps = sum(steps, na.rm = TRUE))
```

Make a histogram of total number of steps taken each day

```{r warning=FALSE, message=FALSE}
plot_steps_complete = total_steps_daily_complete %>%
      ggplot(aes(total_steps))+
      geom_histogram(fill = 'steelblue')+
      labs(title = 'Histogram of step count',
           y = 'Frequency of count',
           x = '')+
      theme_minimal()

ggsave('plot_steps_complete.png',plot_steps_complete)
plot_steps_complete
```
We can see that there are more zeros now.


```{r}
steps_summary_complete = total_steps_daily_complete %>%
      summarise(mean_steps = mean(total_steps),
             median_steps = median(total_steps)) %>%
      as.list() %>%
      unlist

steps_summary_complete
```
What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
rbind(steps_summary_complete, steps_summary)
```

Imputing the data reduced the total mean steps while having similar median steps.

## Are there differences in activity patterns between weekdays and weekends?


Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.


Using lubridate, find the day and identify if weekend or weekday
```{r}
by_week_data = filled_data %>%
      mutate(day = wday(date, label = TRUE)) %>%
      mutate(day_type = factor(ifelse(day == 'Sun' | day == 'Sat', 'weekend', 'weekday')))
```


Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was created using simulated data:

Plot the graph
```{r warning=FALSE, message=FALSE}
weekday_weekend_plot = by_week_data %>%
      group_by(interval, day_type) %>%
      summarise(total_steps = sum(steps, na.rm = TRUE)) %>%
      ggplot(aes(interval, total_steps))+
      geom_line(aes(color = day_type), size = 1.2)+
      facet_grid(day_type ~ ., scales = 'free')+
      labs(title = 'Daily total steps by day type',
           x = '',
           y = 'total steps',
           color = 'day type')+
            theme(legend.position = 'none')

ggsave('weekday_weekend_plot.png',weekday_weekend_plot)

weekday_weekend_plot
```










